Lab 10: Poisson and Negative Binomial Models: Answers

### 2023-01-04 Lab1_R_Review_Answers.Rmd

---
title: '784 Lab 1: R Refresher: Answer Key'
author: "Eric Parajon"
output:
  pdf_document:
    keep_tex: TRUE
  html_document:
    df_print: paged
---

Code from today's lab was partially adapted from labs by Simon Hoellerbauer and Isabel Laterzo.

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, include = TRUE)
```

Today's lab will help you refresh some of what you learned in R in 783. It will also introduce you to some topics in R that you may not be familiar with, but are very useful for this semester.

For a refresher on how to use R Markdown, please see this cheat sheet: https://rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf.

Within your code chunks, comment our your code describing what you are doing in each step. This is an important habit to establish.

#R Review
\medskip
1. First, it is important to save the document you are using in a central folder. Then, set your working directory to that folder. It's also important to clear your environment before working in a new document.
```{r, include = TRUE, eval = FALSE}
getwd() #what is your current working directory?

setwd() #in the parentheses, set the working directory to the folder where
#you save this file. Realistically, this isn't super relevant now because
#you're not reading in external data, but it's a good habit to get into!

rm(list=ls()) #clear global environment
```

\bigskip
2. First, create a vector of ten 5s using the function `rep()`. Second, create a vector that counts from 1 to 10 - do this in two ways, one using `c()` and one using `seq()`. Third, create a vector of length 10 filled with NAs. Finally, uses `cbind()` to bind theses four vectors together to create a 10 x 14 (10 rows, 4 columns) matrix. Remember, if you don't know how to use a function, type ? function name into your console to see the help file (e.g., `?seq()`).

Try binding the vectors together with `rbind()` instead. What happens?

```{r}

vec1 <- rep(5, 10) #vector of ten 5s

vec2 <- c(1:10) #vector counting from 1 to 10

vec3 <- seq(from = 1, to = 10, by = 1) #vector counting from 1 to 10

vec4 <- rep(NA, 10) #vector of 10 NAss

all_vecs <- cbind(vec1, vec2, vec3, vec4) #bind vectorss together using cbind
#we get four columns and ten rows

all_vecs2 <- rbind(vec1, vec2, vec3, vec4) #bind vectors together using rbind
#we get ten columns and four rows

```

\bigskip
3. Set your seed to 425. Simulate 100 draws from a Normal distribution with mean 2 and standard deviation 4, assigning the result to an object.

What proportion of the draws is above 3.1 in our sample? Think about using the functions `sum()` and `length()` to solve this, although there are other ways as well.

What is the expected proportion of draws above 3.1? Hint, look into the function `pnorm()`

Repeat this process, but this time take 10,000 draws from the same Normal distribution. How does our result compare with the expected proportion of draws above 3.1 now?

Thinking back to last semester, what did we just calculate using simulation?

```{r}
set.seed(425)

#simulate 100 draws from N(2, 4)
draws <- rnorm(100, 2, 4)

#calculate proportion of draws above 3.1
sum(draws > 3.1)/length(draws)
#0.39

#using pnorm to retrieve the expected proportion of draws above 1.5
1 - pnorm(3.1, 2, 4)
#just about matches above!

#repeating the above with 100,000 draws
draws2 <- rnorm(10000, 2, 4)
sum(draws2 > 3.1)/length(draws2)
#0.3864, very similar

#We simulated 1 - CDF of N(2, 4) at x = 3.1

```

\bigskip
4. Another important concept in R is a `for` loop. If you need a refresher on these, visit: https://www.r-bloggers.com/2015/12/how-to-write-the-first-for-loop-in-r/.

Using the 100 draws from the Normal distribution from Q3, create a new vector that is 1 when the vector of draws is greater than 3.1 and 0 when it is less than or equal to 3.1. First, do this using a `for` loop with an embedded `if` `else` sequence. What is the sum of this vector?

In addition to loops, `ifelse()` is also an important function. Look at the documentation of this function and repeat the above activity using `ifelse()`.

```{r}
#creating a container (here, a vector) into which we can dump the 1s and 0s. 
#We could do this with c() but it's more efficient to create
#a container of the proper size first 
draws_new <- vector("integer", length(draws))

#for loop that fills the container with 1 when draws > 3.1 and 0 elsewise
for (i in 1:length(draws)){
  if(draws[i] > 3.1){
    draws_new[i] <- 1
  } else draws_new[i] <- 0
}

sum(draws_new) #the number here looks familiar, no?

#repeat above using ifelse
draws_new2 <- ifelse(draws > 3.1, 1, 0)

#checking to make sure we have the same number
sum(draws_new2)
#yay! it is the same.

```

\bigskip
5. Load in the `swiss` dataset. A description for this data can be found here: https://stat.ethz.ch/R-manual/R-patched/library/datasets/html/swiss.html How many variables does it have? How many observations?

The variable `Catholic` is the \% of Catholic individuals, as opposed to Protestant in each province. First look at the summary of this variable. Then, create a new variable `Protestant` that is the percent of Protestant individuals in each province, just going with the assumption that all non-Catholics are Protestant here.

What is the mean percentage of this new variable?

```{r}
#loading in swiss dataset
data(swiss)

#find number of variables and number of observations
ncol(swiss); nrow(swiss)
#six variables, 47 observations

#summary
summary(swiss$Catholic)

#creating protestant variable
swiss$Protestant <- 100 - swiss$Catholic

#summary
summary(swiss$Protestant)

#finding mean of Protestant
mean(swiss$Protestant)
#58.86

```
\bigskip
6. Now, find out what type of object `swiss` is. Then, coerce `swiss` to a list by using `as.list()` and assign it to an objected named `swiss_list`. Look at the object with `View()` What happened? Use the `length()` function on `swiss`. What is the output? Why?

Create a matrix filled with random values that has 4 rows and 2 columns. Use `length()` on this matrix. Why might this be? Use `?` to read the documentation for `length()`.
```{r}
#finding the type of swisss
class(swiss) 
typeof(swiss)

#corece swiss to a list
swiss_list <- as.list(swiss)

#View
View(swiss_list)

#get length of swiss dataset 
length(swiss)

#7 - this is because swiss is a dataframe and a dataframe is really a list of
#same length vectors. This fact helps dataframes mimic matrices.

#this is an example matrix
test_mat <- matrix(c(1, 0, 3, 2, 4, 5, 6, 8), nrow = 4, ncol = 2)
#this is the length
length(test_mat)
#8

#?length()
```
\bigskip
7. Continuing to use the original `swiss` data, use a for loop to calculate the median of each of the variables. Store the result in a *named* vector (i.e., provide a name to each of the values so they are not just a bunch of random numbers with no context!).

Some useful functions here are `vector()`, `seq_along()`, and `names()`. Remember to use the help function `?` for further information about these functions.
```{r}
#creating container for for loop
medians <- vector("double", ncol(swiss))

#for loop that finds the median for each variable (column)
for (i in seq_along(swiss)){
  medians[i] <- median(swiss[, i])
}

#assigning the names of swiss (variable names/column headers) to the median
#vector
names(medians) <- names(swiss)

#print
print(medians)
```
\bigskip
8. Look into the three functions `apply()`, `sapply()`, and `lapply()`. What are the differences between these functions?

Repeat 7, but this time use the three above functions. Using each of these functions, examine what type of object they produce. How are they different from using a `for` loop? How are they different from each other in terms of what they produce? A useful function for examining different object types is `class()`.
```{r}
#examine functions
?apply()
?sapply()
?lapply()

#using apply
medians2 <- apply(swiss, 2, median)
print(medians2)
typeof(medians2)
#it created a named vector

#using lapply
medians3 <- lapply(swiss, median)
print(medians3)
typeof(medians3)
#outputs a named list

#using sapply
medians4 <- sapply(swiss, median)
print(medians4)
typeof(medians4)
#outputs a named vector
```
\bigskip
9. Now we will explore `ggplot2`, which is a great (but sometimess tedious) package for plotting data and results. You will probably grow to have a love/hate relationship with `ggplot2`. For help with this package, reference this cheat sheet: https://rstudio.com/wp-content/uploads/2016/11/ggplot2-cheatsheet-2.1.pdf

Using `ggplot`, create a scatterplot with the `swiss` dataset that plots `Education` against `Examination`. Add axes labels and a title. Does it look like there might be a relationship between these two variables?
```{r}
library(ggplot2)

ggplot(swiss, aes(x = Education, y = Examination)) +
  geom_point() + 
  labs(x = "Education", y = "Examination", title = "Scatterplot") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))

#looks there there could be a positive relationship between Examination and Education 
# as Examination score increases, Education also increases

```

\newpage

\bigskip
10. It is also useful to be able to write your own functions in `R`. Sometimes, `R` doesn't have a function for exactly what you are looking for. Doing so is pretty straightforward. For a quick review, see:  https://swcarpentry.github.io/r-novice-inflammation/02-func-R/

For example, if I wanted to write a function that evaluated the equation \[f(x, y) = xy \] I could write one of the following options:

```{r, include = TRUE}
#example 1
fun_ex1 <- function(params) {
  x <- params[1]
  y <- params[2]

  result <- x*y

  return(result)
}

#evaluate
fun_ex1(c(3,4))

#example 2
fun_ex2 <- function(x, y) {

  result <- x*y

  return(result)
}

#evaluate
fun_ex2(3,4)

```

Now your turn. Write a function that evaluates the following equation:
\[f(x, y, z) = x^2 + 3xy + \sin(z)
\]
Have this function takes a vector as its single argument (see example 1), made up of the three parameters ($x, y, z$). Evaluate the function at $x = 3/4, y = 13, z = 3.4$.

```{r}
#function for the above equation
my_fun <- function(params){
  x <- params[1]
  y <- params[2]
  z <- params[3]

  result <- x^2 + 3*x*y + sin(z)

  return(result)
}

#evaluating the function at x = 3/4, y = 13, z = 3.4
my_fun(c(3/4, 13, 3.4))
```

\newpage
# Math Review:
\bigskip
11. Let's try out some linear algebra. Below are 2 vectors, a and b, and two matrices A and B. Use them for this section.
Here is a link to a linear algebra cheat sheet in R (and other languages): 
https://github.com/scalanlp/breeze/wiki/Linear-Algebra-Cheat-Sheet
```{r, include = TRUE}
#a
a <- c(4, -5, 4, 1)
#b
b <- c(3, 2, -7, 17)

#A
A <- matrix(c(1, -3, 4, 5, 6, 8, 9, -10), ncol = 4)

#B
B <- matrix(c(4:1, 8, 14, -2, 1), ncol = 2)
```

What is the dot product of $\mathbf{a}$ and $\mathbf{b}$? 
Does the result change if we take the dot product of $\mathbf{b}$ and $\mathbf{a}$?

What are the dimensions for the matrix $\mathbf{A}\mathbf{B}$? Is this new matrix invertible?

What are the eigenvectors and eigenvalues of this new matrix?

Verify that the values produced are actually eigenvalues.

```{r}
#dot product
a %*% b
b %*% a

#AB
AB <- A %*% B
solve(AB)
dim(AB)
#yes it is invertible

#eigen vectors and values
eigen_AB <- eigen(AB)

#verifying that they are eigenvalues
sapply(eigen_AB$values, function(x) det(AB - diag(2)*x))

```

### 2023-01-04 Lab 2  Tidyverse Review/Lab2_Tidyverse_Review_Answers.Rmd

---
title: '784 Lab 2: Review of Tidyverse and ggplot Answer key'
author: "Eric Parajon"
output:
  pdf_document:
    keep_tex: TRUE
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, include = FALSE)
```

# Tidyverse Review
\medskip
1. Again, it is important to save the document you are using in a central folder. Then, set your working directory to that folder. It's also important to clear your environment before working in a new document.
```{r, include = TRUE, eval = FALSE}
getwd() #what is your current working directory?

setwd(____) #in the parentheses, set the working directory to the folder where
#you save this file. Realistically, this isn't super relevant now because
#you're not reading in external data, but it's a good habit to get into!

rm(list=ls()) #clear global environment

#The pacman package is required to load/install the additional packages. Uncomment the line below to install the package.
#install.packages("pacman")

#Installing and loading packages for use
pacman::p_load(tidyverse,ggthemes,nycflights13)

```

2.The tidyverse allows us to use the piping operator `%>%`, which is a useful piece of code (base R can do this too with |>) .

It pipes an object (either by itself or created by another function) into the next function in the sequence, as the first argument. This makes the code much easier to read.

For example, we could calculate the standard deviation of the `mpg` variable in the `mtcars` dataset in the following ways:

```{r, include = TRUE, echo = TRUE}
library(tidyverse)

#normal
sd(mtcars$mpg)

#piping
mtcars$mpg %>% 
  sd()

#Does it return the same value?
```
That is a very simple example, of course, and using a pipe in this case does not really gain you much. However, pipes and the myriad of tools in the `tidyverse` can make our work much, much easier. The [`dplyr`](https://dplyr.tidyverse.org/) package in the `tidyverse` allows us to manipulate data much more easily than in base R.

`dplyr` allows us to manipulate our data in a structured way. The key functions are `mutate()`, which creates new variables, `select()`, which allows us to pick variable by characteristics of their names, `filter()`, which allows us to choose rows according to criteria we define, `summarise()`, which allows us to "reduce multiple values down to a single summary", and `arrange()`, which allows us to order the observations according values of 1 or more variables.

RStudio makes useful cheat sheets for their packages, including [`dplyr`](https://www.rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf).

### Star Wars example
```{r}
#1. Star Wars example
data(starwars)
  #Let's make our data into a tibble

#Tibbles are evolution of data.frame
#Inherits nominally, but not strictly!
  #Does not match partial variable names
  #Does not convert character vectors to factors
  #Always returns a tibble when [ is used
  #Allows readable variable names

dplyr::as_tibble(starwars)

  #Now that we have our data imported, we should start exploring it to see exactly what we have

  #the glimpse() function is quite similar to base R's str function
glimpse(starwars)
 #What if we want to pull a specific piece of data?
snip<-slice(starwars, 10:11)

  #What if we want some summary stats on specific variables?
starwars %>%
  group_by(homeworld) %>%
    count() %>% 
arrange(desc(n))

starwars %>%
  group_by(hair_color) %>%
  count() %>% 
  arrange(desc(n))

```

```{r}
#We should start to do some column manipulation now that we have some variables we could be interested in 
d1<-select(starwars, name, species, mass,height,gender,films)

#So now that we have a smaller tibble let's try and do some analysis.
  #What species are the heaviest and which are the lightest?

d1 %>%
  group_by(species) %>%
  summarise(avg = mean(mass, na.rm=TRUE)) %>%
  arrange(desc(avg))
#Hutts are the heaviest!

#Okay great, but what if we want some information that's not explicitly contained in the data? 
  #What if for example, you wanted to know the BMI of a bunch of characters?
  #You can use mutate to create new variables.

Star_BMI<-d1 %>%
  mutate(bmi = mass/((height/100)^2)) %>%
  arrange(desc(bmi)) %>%
  select(name, species, height, mass, bmi)

#Now what if we only want to BMI of Humans characters?

Human_BMI<-Star_BMI %>% 
  filter(species=="Human")

#What if we wanted to create a new variable that combined a characters name with where they were born?
#This is easy using tidyr (another package in tidyverse suite)
starwars %>%
  unite(name_homeworld, name, homeworld) %>% 
  select(name_homeworld)

```
## Joins

```{r}

# Load data 1
(dates <- read_csv("https://bit.ly/DM22_dates"))

# Load data 2
(works <- read_csv("https://bit.ly/DM22_works"))

#Inner join
#With an inner_join(x, y), only observations that appear in both x and y datasets are returned

# Do inner join
dates_works <- inner_join(dates, works) 
(dates_works)

#Left & Right Join
#A left_join(x, y) call returns all observations in x, and drops observations that only appear in y
#Variables in y that do not exist in x are appended to x in the returned tibble.

#The same is true for a right_join(x, y), but with the roles of x and y reversed.

dates_someworks <- left_join(dates, works) 
(dates_someworks)

dates_allworks <- right_join(dates, works) 
print(dates_allworks, width=70)

#Full join
#Finally, a full_join(x, y) returns all observations and all variables in either in x or in y (filled with NAs)

alldates_allworks <- full_join(dates, works) 
print(alldates_allworks, width=70)

```

## Pivoting
Pivoting: Wide -> Long
With pivot_longer(), you can take values of a variable that are stored as variable names pivot them onto the cells of a single (implicit) variable column

```{r}
(relig_income)
#This data set has counts based on a survey which (among other things) asked people about their religion and annual income:

relig_income_long<-relig_income %>% 
  pivot_longer(!religion, names_to = "income", values_to = "count")

#The first argument is the dataset to reshape, relig_income.

#The second argument describes which columns need to be reshaped. In this case, it’s every column apart from religion.

#The names_to gives the name of the variable that will be created from the data stored in the column names, i.e. income.

#The values_to gives the name of the variable that will be created from the data stored in the cell value, i.e. count.

```
Pivoting: Long -> Wide

With pivot_wider(), we move in the opposite direction: your units are distributed over multiple rows, so pivot until you have a single case per row of your tibble

```{r}
(fish_encounters) 
#This dataset, contributed by Myfanwy Johnston, describes when fish swimming down a river are detected by automatic monitoring stations:

fish_encounters_wide<- fish_encounters %>% 
          pivot_wider(names_from = station, values_from = seen)

#In this case the NAs mean no fish was recorded so we can convert that to zero like this

fish_encounters %>% pivot_wider(
  names_from = station, 
  values_from = seen,
  values_fill = 0
)

```
#Part 2: ggplot

`ggplot` is very important, and requires some time and patience to get to know well. We won't have time to dive fully into it, but we will begin reviewing the following code so you become more familiar. R has its own built-in plot function, but ggplot is far more flexible and creates far more descriptive (and thus useful!) plots.

For a comprehensive 'cheat sheet', see: https://www.maths.usyd.edu.au/u/UG/SM/STAT3022/r/current/Misc/data-visualization-2.1.pdf

```{r}
#we have to load the package first
library(ggplot2)
```

To begin, we can use the 'qplot' or 'ggplot' commands. I use ggplot. It's more customization.

```{r}
data(mtcars)

qplot(x = disp, y = mpg, data = mtcars, color = cyl, geom = "point") 
#x and y are independent + dependent variables, 
#data can be specified (or use mtcars$var),
#color distinguishes a characteristic over which to map color
#and geom is a plot type

ggplot(data = mtcars, aes(x = disp, y = mpg)) + geom_point(aes(color = cyl))
#Note the similarities above. 
#ggplot is more flexible in that you map to different `aesthetics'- 
#hence why we use 'aes.' Plenty of examples below to dig into this more! 
```

What are some ways we can make this plot clearer?

```{r}
ggplot(data = mtcars, aes(x = disp, y = mpg)) + #we add 'layers' with +
  geom_point(aes(colour = cyl), size = 2) + #point=scatterplot
  labs(title = "Car Stuff", x = "Displacement (cc)", y = "Miles per Gallon")+ 
  #label x and y axes, give the plot a main title. Note the quotes!
  scale_colour_continuous(name = "Cylinders")+ #label the scale
  theme_bw() #make the plot background blank w/grid lines
```

You can also assign this plot to an object and save that object as a .pdf file.

```{r, message=FALSE, warning=FALSE}
carsplot <- ggplot(data = mtcars, aes(x = disp, y = mpg)) + 
  geom_point(aes(colour = cyl), size = 2) +
  labs(title = "Car Stuff", x = "Displacement (cc)", y = "Miles per Gallon") +
  scale_colour_continuous(name = "Cylinders")+
  theme_bw()

#One method for saving plots:
#pdf("Plot1.pdf")
#carsplot
#dev.off()

#Where do these plots go?!

#Unique to ggplot:
#ggsave("Plot1a.pdf") #can edit image sizes as well
```
## Themes
As you've hopefully seen ggplot is highly customization. One way to standardize your graphs and avoid repeating your code to control ancillary parameters (e.g. color palettes, axes labels, etc.) is to use a ggplot() theme that includes bundles of code. This is what we were doing using theme_bw() above.

You can also create custom themes yourself!

```{r}
#Here is a theme I use often

plot_theme = theme(text = element_text(size=11),  #set text size to 11
                   panel.grid.minor = element_blank(), #these next two lines set the grid lines to null
                   panel.grid.major = element_blank(),
                   panel.background = element_blank(),#these lines make the background of the plot blank
                   plot.background = element_rect(fill = "transparent",colour = NA),
                   axis.text = element_text(color="black"), #setting axis text to black 
                   axis.line = element_line(colour = "black"),#setting axis lines to black 
                   strip.text.y = element_text(angle = 0), axis.title.y = element_text(angle=90, hjust=.5, vjust=.5),#adjusting the angle of text
                   axis.title = element_text(lineheight=1.1), #increasing height of axis title
                   legend.position = "bottom",  #Placing the legend at the bottom of the figure
                   legend.key = element_rect(colour = "transparent",  fill = "transparent") #setting legend to transparent
)

#Using the theme
ggplot(data = mtcars, aes(x = disp, y = mpg)) + 
  geom_point(aes(colour = cyl), size = 2) +
  labs(title = "Car Stuff", x = "Displacement (cc)", y = "Miles per Gallon") +
  scale_colour_continuous(name = "Cylinders")+
  plot_theme

#As you can see there are tons of options that you can set yourself. Alternatively you can use existing gplot_themes. Here are some of the most popular
#bw
ggplot(data = mtcars, aes(x = disp, y = mpg)) + 
  geom_point(aes(colour = cyl), size = 2) +
  labs(title = "Car Stuff", x = "Displacement (cc)", y = "Miles per Gallon") +
  scale_colour_continuous(name = "Cylinders")+
  theme_bw()

#minimal
ggplot(data = mtcars, aes(x = disp, y = mpg)) + 
  geom_point(aes(colour = cyl), size = 2) +
  labs(title = "Car Stuff", x = "Displacement (cc)", y = "Miles per Gallon") +
  scale_colour_continuous(name = "Cylinders")+
  theme_minimal()

#classic
ggplot(data = mtcars, aes(x = disp, y = mpg)) + 
  geom_point(aes(colour = cyl), size = 2) +
  labs(title = "Car Stuff", x = "Displacement (cc)", y = "Miles per Gallon") +
  scale_colour_continuous(name = "Cylinders")+
  theme_classic()

#If you want even more themes you can use the ggthemes library!
#install.packages("ggthemes")
library(ggthemes)

#WSJ
ggplot(data = mtcars, aes(x = disp, y = mpg)) + 
  geom_point(aes(colour = cyl), size = 2) +
  labs(title = "Car Stuff", x = "Displacement (cc)", y = "Miles per Gallon") +
  scale_colour_continuous(name = "Cylinders")+
  theme_wsj()

#538
ggplot(data = mtcars, aes(x = disp, y = mpg)) + 
  geom_point(aes(colour = cyl), size = 2) +
  labs(title = "Car Stuff", x = "Displacement (cc)", y = "Miles per Gallon") +
  scale_colour_continuous(name = "Cylinders")+
  theme_fivethirtyeight()

#The Economist
ggplot(data = mtcars, aes(x = disp, y = mpg)) + 
  geom_point(aes(colour = cyl), size = 2) +
  labs(title = "Car Stuff", x = "Displacement (cc)", y = "Miles per Gallon") +
  scale_colour_continuous(name = "Cylinders")+
  theme_economist()

```
## Facets
Using facets lets you create multi-panel plots based off a given variable.
There are two different ggplot functions to use to facet. 
```{r}
#1. facet_wrap() "wraps" a 1d ribbon of panels into 2d. Essentially, creating a square grid of panels, with rows and columns defined by combinations of two categorical variables (even if some combinations of faceting values do not exist in data). This is most useful when you want to facet by a single variable (or two)

carsplot_facet <- ggplot(data = mtcars, aes(x = disp, y = mpg)) + 
  geom_point(aes(colour = cyl), size = 2) +
  labs(title = "Car Stuff", x = "Displacement (cc)", y = "Miles per Gallon") +
  scale_colour_continuous(name = "Cylinders")+
  theme_bw()+
  facet_wrap(~ am)

#2. facet_grid() forms a matrix of panels defined by row and column faceting variables, creating panels based on observed combinations of values in faceting variables

#facet_grid(.~variable) will return facets equal to the levels of variable distributed horizontally
#facet_grid(variable~.) will return facets equal to the levels of variable distributed vertically.

carsplot_facet_2 <- ggplot(data = mtcars, aes(x = disp, y = mpg)) + 
  geom_point(aes(colour = cyl), size = 2) +
  labs(title = "Car Stuff", x = "Displacement (cc)", y = "Miles per Gallon") +
  scale_colour_continuous(name = "Cylinders")+
  theme_bw()+
  facet_grid(vs ~ am)

carsplot_facet_3 <- ggplot(data = mtcars, aes(x = disp, y = mpg)) + 
  geom_point(aes(colour = cyl), size = 2) +
  labs(title = "Car Stuff", x = "Displacement (cc)", y = "Miles per Gallon") +
  scale_colour_continuous(name = "Cylinders")+
  theme_bw()+
  facet_grid(am ~ .)

carsplot_facet_4 <- ggplot(data = mtcars, aes(x = disp, y = mpg)) + 
  geom_point(aes(colour = cyl), size = 2) +
  labs(title = "Car Stuff", x = "Displacement (cc)", y = "Miles per Gallon") +
  scale_colour_continuous(name = "Cylinders")+
  theme_bw()+
  facet_grid(. ~ am)
```

What are other types of common geoms?

geom_histogram(): plots a histogram of single value; usually needs modification to number of bins.

geom_density(): density plot of single variable.

geom_bar(): vertical barplot by categorical variable

geom_point(): typical two-way scatterplot

geom_smooth(): simple line interpolator

#Part 3: Practice
## Tidyverse practice

Use the data below to first split the complete dataset into individual planes and then summarise each plane 
by counting the number of flights (count = n()) and computing the mean distance flown.

Next, determine what city is flown to the most often from NYC.

Finally, create a plot to create a barplot visualizing if flights are on time or delayed (defined as later than 5 minutes) by month. Additionally, ensure your plot has meaningful labels.

```{r}
#First, we need to install our data. This dataset includes airline on-time data for all flights departing NYC in 2013. 
#install.packages("nycflights13")

library(nycflights13)

data("flights")
dplyr::as_tibble(flights)

```

```{r}
by_tailnum <- group_by(flights, tailnum) %>% 
  summarise(count = n(), dist = mean(distance, na.rm = TRUE))

```

```{r}
#What city is flown to the most often from NYC?

destinations <- group_by(flights, dest) %>%
  summarise(planes = n_distinct(tailnum),
            flights = n())%>%
  arrange(desc(flights))

#calling to view top 10 
d1<-head(destinations,10)

```

```{r}
#Barplot
flight_delays<-flights %>% 
  mutate(dep_type = ifelse(dep_delay < 5, "on time", "delayed")) %>% 
   dplyr::group_by(dep_type,month) %>%
    dplyr::summarise(n = n()) %>% 
    filter(dep_type!="")

plot1<-ggplot(data=flight_delays, aes(x = month,y=n, fill = dep_type)) +
  geom_bar(stat="identity")+
  labs(main="Frequency of On Time or Delayed Arrivals by Month", y="Count", fill="Type")

```

### 2023-01-04 Lab 10 PoissonNBmodels/Lab10_PoissonNBmodels_AnswerKey.Rmd
Add files via upload

---
title: "Lab 10: Poisson and Negative Binomial Models"
author: "Eric Parajon"
output: pdf_document
---

@@

Load necessary packages:
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, messsage = FALSE)
rm(list=ls())

#Load in packages
pacman::p_load(tidyverse,
               haven,#for importing data
               labelled,
               MASS,
               AER,
               foreign)
```

### Poisson and Negative Binomial Models

@@

(These data are available on Sakai.)

```{r, message=F, warning=F, include=F}
count <- read_dta("Maoz.dta")

```

We'll be using the variable "nowars20" as our DV - this counts the number of wars in the international system. Let's plot a histogram:

```{r, message=F, warning=F}
ggplot(data = count , aes(x= nowars20 )) +
  geom_histogram(bins=25, fill="turquoise") +
  labs(x="Number of MIDS") +
  theme_bw()
```

@@

```{r}
m1 <- glm(nowars20 ~ lgtrdnpi2 + clqdemreg,
          data = count,
          family = "poisson")
summary(m1)
```
Note that this tells us "Dispersion parameter for poisson family taken to be 1."

@@

One approach is to plot the mean of the data against the estimated variance. Try doing this by using the `fitted()` function below (look up what it is!), and then following the code.

```{r, message=F, warning=F}
fits <- fitted(m1) #fitted()

@@

```{r, message=F, warning=F}
dispersiontest(m1)
```
So, our ALTERNATIVE hypothesis is that the true dispersion is greater than 1. What is our p-value? Do we reject the null? (What is the null?)

@@

YES! Looks like we might be dealing with some overdispersion. What can we do to correct this? Let's try a new model - a negative binomial model. While Poisson assumes that the mean and variance are the same, the Negative Binomial model better fits data that shows variation which is greater than the mean. The NB model has an additional parameter compared to the Poisson, and this parameter allows us to adjust for more dispersion in the data. In fact, the Poisson model is a special case of the NB.

@@

m2<- glm.nb(nowars20 ~ lgtrdnpi2 + clqdemreg, 
            data = count)
summary(m2)

@@

```
What happened to the coefficients?

@@

```{r}
#compare models using AIC
#which is better?

AIC(m1)
AIC(m2)

#AIC in m2 is lower
```

Let's also compare these models using a rootogram. We're going to install a package here from a repo which makes it easy to do this. Rootograms help us understand the predictive accuracy of a model based on actual data. It will show us the expected counts given the model vs. the observed counts. It does so by square rooting the counts, which allows us to more easily examine departures from our expectations even at smaller frequencies.

```{r}
install.packages("countreg", repos="http://R-Forge.R-project.org")
library(countreg)
library(cowplot)

@@

## Predicted Counts

@@

#Set a number of simulations. 
m2_sims <-1000

#Create our matrix of simulated data- we'll limit our dataset to do so.
sims_mat <- mvrnorm(m2_sims, Betas, vcv)
dim(sims_mat)

@@

#Fill in our predicted probabilities with a for() loop! 
for(i in 1:m2_sims){
  pc_sims[i, ] <- exp(temp_data%*%sims_mat[i, ]) 
}

@@

To do on your own:

Below, try using ggplot to plot this information we just generated. Follow these guidelines:
- X axis: Average number of Democratic Cliques
- Y axis: Predicted count of Intl. Wars
- Include a geom_line() for Dem_Cliques
- Include a geom_ribbon() with CIs (80% confidence interval)
- Include a legend which specifies what the line is and what the CIs are, try putting that legend on the bottom

@@

pcplot <- ggplot(_____)

pcplot

```

What do you find?

>>>>

---
title: "Lab 10: Poisson and Negative Binomial Models: Answers"
author: "TA: Eric Parajon"
output: pdf_document
---

@@

Load necessary packages:
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, messsage = FALSE)
rm(list=ls()) 
#Load in packages
pacman::p_load(tidyverse,
               haven,#for importing data
               labelled,
               MASS,
               AER,
               foreign)
```

### Poisson

@@

*You'll need to change the directory to bring in the file.*

```{r, message=F, warning=F, include=F}
count <- read_dta("Maoz.dta")
#count2 <-na.omit(count) # <- don't do this with your data unless you are absolutely
#sure that you want to drop all nas 
```

We'll be using the variable "nowars20" as our DV - this counts the number of wars in the international system. Here's our histogram:

```{r, message=F, warning=F}
ggplot(data=count, aes(x=nowars20)) +
  geom_histogram(bins=25, fill="turquoise") +
  labs(x="Number of MIDS") +
  theme_bw()
```

@@

```{r}
m1 <- glm(data = count, nowars20 ~ lgtrdnpi2 + clqdemreg, family="poisson")
summary(m1)
```
Note that this tells us "Dispersion parameter for poisson family taken to be 1."

@@

One approach is to plot the mean of the data against the estimated variance.

```{r, message=F, warning=F}
fits <- fitted(m1) #what does fitted do, do you think?

@@

```{r, message=F, warning=F}
dispersiontest(m1)
```
So, our ALTERNATIVE hypothessis is that the true dispersion is greater than 1. What is our p-value? So what is supported - null or alternative hyptohesis? Do we have overdispersion?

@@

YES! Looks like we might be dealing with some overdispersion. What can we do to correct this? Let's try a new model - a negative binomial model. While Poisson assumes that the mean and variance are the same, the Negative Binomial model better fits data shows variation which is greater than the mean. The NB model has an additional parameter compared to the Poisson, and this parameter allows us to adjust for more dispersion in the data. In fact, the Poisson model is a special case of the NB.

@@

m2<- glm.nb(data = count, nowars20~lgtrdnpi2+clqdemreg)
summary(m2)

@@

```
What happened to the coefficients? What happened to the standard errors?

@@

```{r}
AIC(m1)
AIC(m2)
#which is better?
```

Let's also compare these models using a rootogram:

```{r}
#install.packages("countreg", repos="http://R-Forge.R-project.org")
library(countreg)
library(cowplot)

@@

#Predicted Counts

@@

#Set a number of simulations. 
m2_sims<-1000

#Create our matrix of simulated data- we'll limit our dataset to do so.
sims_mat <- mvrnorm(m2_sims, Betas, vcv) #asymptotic sampling distribution of the coefs is the same as in any other GLM
dim(sims_mat)

@@

#Fill in our expected counts with a for() loop
for(i in 1:m2_sims){
  pc_sims[i, ] <- exp(temp_data%*%sims_mat[i, ]) 
}

@@

@@

pc_plot <- ggplot(data=plot_data) + 
  labs(x = "Average Number of Democratic Cliques", 
       y = "Predicted Count of Intl. Wars") +
  geom_line(aes(x = Dem_Cliques, y = med, 
                color = "Number of Interstate Wars"), size = 1) +
  geom_ribbon(aes(x = Dem_Cliques, ymin = low, ymax = high, 
                  fill = "80% Confidence Interval (Simulation)"), alpha = .5) +
  theme_bw() +
  theme(legend.title = element_text(), 
        legend.position = "bottom", panel.grid.minor.x=element_blank(),
        panel.grid.minor.y = element_blank()) +
  scale_fill_manual(values = "steelblue3", name = "")+
  scale_color_manual(values = "black", name = "")
pc_plot
```
